---
title: "Import and representation of ProteinGym data"
author: 
    - name: Tram Nguyen
      affiliation: Department of Biomedical Informatics, Harvard Medical School
      email: Tram_Nguyen@hms.harvard.edu
    - name: Pascal Notin
      affiliation: Department of Systems Biology, Harvard Medical School
    - name: Aaron W Kollasch
      affiliation: Department of Systems Biology, Harvard Medical School
    - name: Debora Marks
      affiliation: Department of Systems Biology, Harvard Medical School
    - name: Ludwig Geistlinger
      affiliation: Department of Biomedical Informatics, Harvard Medical School
package: ProteinGymR
output:
    BiocStyle::html_document:
      self_contained: yes 
      toc: true
      toc_float: true
      toc_depth: 2
      code_folding: show
date: "April 14, 2025"
vignette: >
    %\VignetteIndexEntry{Data access and representation}
    %\VignetteEncoding{UTF-8}
    %\VignetteEngine{knitr::rmarkdown}
editor_options: 
    markdown: 
      wrap: 80
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    crop = NULL,
    message = FALSE
)
```

# Installation

Install the package using Bioconductor. Start R and enter:

```{r, eval = FALSE}
if(!requireNamespace("BiocManager", quietly = TRUE))
        install.packages("BiocManager")
BiocManager::install("ProteinGymR")
```

# Setup

Now, load the package and dependencies used in the vignette.

```{r, message = FALSE}
library(ProteinGymR)
library(tidyr)
library(dplyr)
library(stringr)
library(ggplot2)
```

[ProteinGym]: https://proteingym.org/
[pg_publication]: https://proceedings.neurips.cc/paper_files/paper/2023/hash/cac723e5ff29f65e3fcbb0739ae91bee-Abstract-Datasets_and_Benchmarks.html
[zenodo]: https://zenodo.org/records/14997691
[dms_paper]: https://www.nature.com/articles/nmeth.3027
[physiochem]: https://biology.stackexchange.com/questions/105321/arrangement-of-amino-acids-in-the-protein-alphabet
[boucher]: https://pubmed.ncbi.nlm.nih.gov/27010590/

# Introduction

Predicting the effects of mutations in proteins is critical to many
applications, from understanding genetic disease to designing novel proteins
to address our most pressing challenges in climate, agriculture and
healthcare. Despite an increase in machine learning-based protein modeling
methods, assessing the effectiveness of these models is problematic due to
the use of distinct, often contrived, experimental datasets and variable
performance across different protein families.

[ProteinGym][ProteinGym] is a large-scale and holistic set of benchmarks 
specifically designed for protein fitness prediction and design curated by
([Notin et al. 2023][pg_publication]). It encompasses both a broad collection 
of over 250 standardized deep mutational scanning (DMS) assays, spanning 
millions of mutated sequences, as well as curated clinical datasets providing 
high-quality expert annotations about mutation effects. Furthermore, ProteinGym 
reports the performance of a diverse set of over 70 high-performing models from 
various subfields (eg., mutation effects, inverse folding) into a unified 
benchmark.

ProteinGym datasets are openly available as a community resource both on
[Zenodo][zenodo] and the official [ProteinGym website][ProteinGym]. 
ProteinGym v1.1 was released with Bioconductor 3.20 and v1.2 with 
Bioconductor 3.21.


# Data

The `ProteinGymR` package provides the following analysis-ready datasets from 
ProteinGym:

1.  DMS assay scores from 217 assays measuring the impact of all possible 
amino acid substitutions across 186 proteins. The data is provided with 
`dms_substitutions()`.

2.  AlphaMissense pathogenicity scores for ~1.6 M substitutions in the
ProteinGym DMS data. The data is provided with `am_scores()`.

3. Reference file containing metadata associated with the 217 DMS assays.

4. Five model performance metrics ("AUC", "MCC", "NDCG", "Spearman", 
"Top_recall") for 79 models across 217 assays (as of Bioc 3.21) calculated on 
DMS substitutions in a zero-shot setting. The data is provided with 
`zeroshot_DMS_metrics()`.

5. Model scores on the DMS substitutions for 79 models in the zero-shot setting.
Load in with `zeroshot_substitutions()`.


6. Two model performance metrics ("Spearman", and "MSE") for 12 models across 
217 assays (as of Bioc 3.21) calculated on DMS substitutions in a 
semi-supervised setting. Load in this data with `supervised_metrics()`.

7. Model scores on the DMS substitutions for 11 semi-supervised models with 3 
folding schemes: contiguous, modulo, and random. Loaded in with 
`supervised_substitutions()` and by changing the "fold_scheme" argument, 
respectively.

8. PDB files for 197 protein structures, to be used in the `plot_structure()` 
3D visualization function.


# Import ProteinGym data

ProteinGym data is accessed through ExperimentHub.


## Load and explore the DMS data

Deep mutational scanning is an experimental technique that provides 
comprehensive data on the functional effects of all possible single mutations 
in a protein ([Fowler & Fields 2014][dms_paper]). For each position in a 
protein, the amino acid residue is mutated and the fitness effects are recorded. 
While most mutations tend to be deleterious, some can enhance protein activity. 
In addition to analyzing single mutations, this method can also examine the 
effects of multiple mutations, yielding insights into protein structure and 
function. Overall, DMS scores provide a detailed map of how changes in a 
protein's sequence affect its function, offering valuable yet complex insights 
for researchers studying protein biology.

Datasets in `ProteinGymR` can be easily loaded with built-in functions.

```{r import dms}
dms_data <- dms_substitutions()
```

View the DMS study names for the first 6 assays.
```{r view studies}
head(names(dms_data))
```

View an example of one DMS assay.
```{r view assay}
head(dms_data[[1]])
```

For each DMS assay, the columns show the UniProt protein identifier, the DMS 
experiment assay identifier, the mutant at a given protein position, the mutated
protein sequence, the recorded DMS score, and a binary DMS score bin 
categorizing whether the mutation has an affect on fitness (1) or not (0). For 
more details, access the function documentation with `?dms_substitutions()` and
the reference publication from Notin et al. 2023.

To access the metadata associated with each DMS assay, we can load in the 
reference table. Do this by querying all datasets on ExperimentHub affilitated
with "ProteinGymR".

```{r queryEH}
eh <- ExperimentHub::ExperimentHub()
AnnotationHub::query(eh, "ProteinGymR")

dms_metadata <- eh[["EH9607"]]
names(dms_metadata)
```

There are 45 columns representing metadata for DMS assays. For more information 
about the information, see the [ProteinGym publication][pg_publication].


# Benchmarking across models

We will now use the built-in function `benchmark_models()` to compare 
performance across several variant effect prediction models calculated on the 
217 DMS assays in the zero-shot setting. This function takes in one of the five 
available metrics, and compares up to 5 models of the 62 available.

In the zero-shot setting, experimental phenotypical measurements from
a given assay are predicted without having access to any ground-truth labels at 
training time. Robust zero-shot performance is particularly informative when 
labels are subject to several biases or scarcely available (e.g., labels for 
rare genetic pathologies).

Model performance was evaluted across 5 metrics:

1. Spearman's rank correlation coefficient (primary metric)
2. Area Under the ROC Curve (AUC)
3. Matthews Correlation Coefficient (MCC) for bimodal DMS measurements
4. Normalized Discounted Cumulative Gains (NDCG) for identifying the most 
functional protein variants
5. Top K Recall (top 10% of DMS values)

To avoid placing too much weight on properties with many assays 
(e.g., thermostability), these metrics were first calculated within groups 
of assays that measure similar functions. The final value of the metric
is then the average of these averages, giving each functional group equal 
weight. The final values are referred to as the ‘corrected average’.

Due to the often non-linear relationship between protein function and 
organism fitness ([Boucher et al., 2016][boucher]), the Spearman’s rank 
correlation coefficient is the most generally appropriate 
metric for model performance on experimental measurements. However, in 
situations where DMS measurements exhibit a bimodal profile, rank correlations 
may not be the optimal choice. Therefore, additional metrics are also provided, 
such as the Area Under the ROC Curve (AUC) and the Matthews Correlation 
Coefficient (MCC), which compare model scores with binarized experimental 
measurements. Furthermore, for certain goals (e.g., optimizing functional 
properties of designed proteins), it is more important that a model is able to 
correctly identify the most functional protein variants, rather than properly 
capture the overall distribution of all assayed variants. Thus, we also 
calculate the Normalized Discounted Cumulative Gains (NDCG), which up-weights a 
model if it gives its highest scores to sequences with the highest DMS value. 
Finally, we also calculate Top K Recall, where we select K to be the top 10% 
of DMS values.

To view all available models, use the function: `available_models()`

```{r, available_models}
available_models()
```

Plot the AUC metric for 5 models.

```{r, warning=FALSE, fig.wide = TRUE}
benchmark_models(metric = "AUC", 
    models = c("GEMME", "CARP_600K", "ESM1b", "VESPA", "ProtGPT2"))
```

Based on the AUC metric of evaluation, GEMME performed the best while of the 5 
selected models. If the `metric` argument is not defined, the default used is 
a Spearman correlation. For more information about the models and metrics, see 
the function documentation `?benchmark_models()`.


# Reference

Notin, P., Kollasch, A., Ritter, D., van Niekerk, L., Paul, S., Spinner, H., 
Rollins, N., Shaw, A., Orenbuch, R., Weitzman, R., Frazer, J., Dias, M., 
Franceschi, D., Gal, Y., & Marks, D. (2023). ProteinGym: Large-Scale 
Benchmarks for Protein Fitness Prediction and Design. In A. Oh, T. Neumann, 
A. Globerson, K. Saenko, M. Hardt, & S. Levine (Eds.), Advances in Neural 
Information Processing Systems (Vol. 36, pp. 64331-64379). 
Curran Associates, Inc.

Fowler, D., Fields, S. Deep mutational scanning: a new style of protein science. 
Nat Methods 11, 801–807 (2014). doi: 10.1038/nmeth.3027.

Boucher JI, Bolon DN, Tawfik DS. Quantifying and understanding the fitness 
effects of protein mutations: Laboratory versus nature. Protein Sci. 2016 Jul;
25(7):1219-26. doi: 10.1002/pro.2928.


# Session Info
```{r, sesh info}
sessionInfo()
```