---
title: "3D Protein Structure Exploration"
author: 
    - name: Tram Nguyen
      affiliation: Department of Biomedical Informatics, Harvard Medical School
      email: Tram_Nguyen@hms.harvard.edu
    - name: Pascal Notin
      affiliation: Department of Systems Biology, Harvard Medical School
    - name: Aaron W Kollasch
      affiliation: Department of Systems Biology, Harvard Medical School
    - name: Debora Marks
      affiliation: Department of Systems Biology, Harvard Medical School
    - name: Ludwig Geistlinger
      affiliation: Department of Biomedical Informatics, Harvard Medical School
package: ProteinGymR
output:
    BiocStyle::html_document:
      self_contained: yes 
      toc: true
      toc_float: true
      toc_depth: 2
      code_folding: show
date: "`r doc_date()`"
vignette: >
    %\VignetteIndexEntry{3D Protein Structure}
    %\VignetteEncoding{UTF-8}
    %\VignetteEngine{knitr::rmarkdown}
editor_options: 
    markdown: 
      wrap: 80
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 3D protein structure

This vignette demonstrates how to explore and visualize DMS scores on a 3D 
protein structure using the package r3dmol under the hood. The function requires
a file path to a PBD file contain protein structure coordinates and a DMS assay
to aggregate DMS scores that will be projected onto the 3D structure.

```{r, message=FALSE, warning=FALSE, execute = FALSE}
library(ProteinGymR)
```

By default, if no `dms_data` argument is provided, the DMS subsitutions from 
`ProteinGymR::dms_substitutions()` are loaded in. Alternatively, a user can 
specify their own dms list to project. The user may also specify what
aggregation method to use for calucating the statistic at each residue position.
By default, the mean DMS score is calculated for each position. See the function
documentation for details: `?plot_structure()`

First, let's plot using all default settings. The only required arguments are
the `assay_name` and the `pdb_file` path.

Importantly, the function normalizes the aggregated DMS scores so that a score
of zero will always be represented as white and the range of scores will be
scaled relatively to -1 and 1 in that assay, with 1 being the maximum 
aggregated DMS score within that assay region.

```{r}
plot_structure(assay_name = "ACE2_HUMAN_Chan_2020", 
    pdb_file = "~/ProteinGym_data/ProteinGym_v1.1/ProteinGym_AF2_structures/ACE2_HUMAN.pdb")
```

<br>

In this example, we are plotting the 3D structure of the ACE2_HUMAN protein and
overlaying the mean DMS score across all mutants in a given position. Chan et 
al. 2020 who generated the DMS assay data only experimentally assessed a subset
of the entire ACE2_HUMAN protein. By default the function only displayed the 
regions where there is information available in the assay. Red colors represent
more pathogenic (lower DMS scores) and blue colors show more benign positions
(higher DMS scores). Regions that appear white indicate closer to no change 
before and after the DMS perturbation.

If we want to visualize the complete protein structure, but simple grey out the
regions we do not have DMS assay information for, we can set the argument
`full_structure == TRUE`

```{r}
plot_structure(assay_name = "ACE2_HUMAN_Chan_2020", 
    pdb_file = "~/ProteinGym_data/ProteinGym_v1.1/ProteinGym_AF2_structures/ACE2_HUMAN.pdb",
    full_structure = TRUE)
```

<br>

Now we can more clearly see the entire protein structure for ACE2_HUMAN. 

Some assays extensively assessed nearly every position of the complete protein, 
for example: the C6KNH7_9INFA protein from Lee et al. 2018. Let's visualize 
this protein and set the aggregation method to view the minimum DMS score across
all mutants at each position by setting `aggregation_fun = min`.

```{r}
plot_structure(assay_name = "C6KNH7_9INFA_Lee_2018", 
    pdb_file = "~/ProteinGym_data/ProteinGym_v1.1/ProteinGym_AF2_structures/C6KNH7_9INFA.pdb",
    aggregate_fun = min)
```

<br>

For this protein, it looks like 

```{r}
dms_data <- dms_substitutions()
df <- dms_data[["C6KNH7_9INFA_Lee_2018"]]
hist(df$DMS_score)
```

We can see that for this DMS assay, most values of the minimum DMS score at each
position are negative and are thus represented as red.

<br>

Finally, the user can also zoom into a specific region of the protein by
defining residue position coordinates with `start_pos` and `end_pos`. 

Let's just visualize the mean DMS scores of the protein from position 10:50.

```{r}
plot_structure(assay_name = "ACE2_HUMAN_Chan_2020", 
    pdb_file = "~/ProteinGym_data/ProteinGym_v1.1/ProteinGym_AF2_structures/ACE2_HUMAN.pdb",
    start_pos = 10,
    end_pos = 50)
```


## Session Info
```{r}
sessionInfo()
```

