---
title: "Visualization and Exploration"
author: 
    - name: Tram Nguyen
      affiliation: Department of Biomedical Informatics, Harvard Medical School
      email: Tram_Nguyen@hms.harvard.edu
    - name: Pascal Notin
      affiliation: Department of Systems Biology, Harvard Medical School
    - name: Aaron W Kollasch
      affiliation: Department of Systems Biology, Harvard Medical School
    - name: Debora Marks
      affiliation: Department of Systems Biology, Harvard Medical School
    - name: Ludwig Geistlinger
      affiliation: Department of Biomedical Informatics, Harvard Medical School
package: ProteinGymR
output:
    BiocStyle::html_document:
      self_contained: yes 
      toc: true
      toc_float: true
      toc_depth: 2
      code_folding: show
date: "`r doc_date()`"
vignette: >
    %\VignetteIndexEntry{Visualization and Exploration"}
    %\VignetteEncoding{UTF-8}
    %\VignetteEngine{knitr::rmarkdown}
editor_options: 
    markdown: 
      wrap: 80
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

[CH]: https://bioconductor.org/packages/release/bioc/html/ComplexHeatmap.html
[r3d]: https://cran.r-project.org/web/packages/r3dmol/vignettes/r3dmol.html
[pg_publication]: https://proceedings.neurips.cc/paper_files/paper/2023/hash/cac723e5ff29f65e3fcbb0739ae91bee-Abstract-Datasets_and_Benchmarks.html

# Setup

```{r, message=FALSE, warning=FALSE}
library(ProteinGymR)
library(BiocFileCache)
```


# Introduction 

This vignette demonstrates how to explore and visualize DMS and model scores 
from the ProteinGym database v1.2. Specifically, it walks through the 
functionality to generate heatmaps of all pairwise DMS substitution mutants and 
projects these scores onto 3D protein structures. `ProteinGymR` uses 
functionality from the Bioconductor package [ComplexHeatmap][CH] and 
[r3dmol][r3d] from CRAN under the hood to generate the heatmaps and 3D protein 
structures, respectively. 


# Visualize DMS scores along a protein

Explore the "ACE2_HUMAN" assay from Chan et al. 2020 and create a heatmap of 
the DMS scores with `plot_dms_heatmap()`. If the argument `dms_data` is not 
specified, the default will load the most recent DMS substitution data from 
ProteinGym with `ProteinGymR::dms_substitutions()`. This function only requires
a specific assay name. To obtain all assay names, run: 
`names(dms_substitutions())`. By default, the function also plots the full range
of positions where DMS data is available for this assay. To plot a specific 
region of interest, use the arguments `start_pos()` and `end_pos()` which takes
in an integer for the first and last residue position to plot in the protein.

```{r ACE2 default heatmap, fig.wide = TRUE}
ace2_dms <- plot_dms_heatmap(assay_name = "ACE2_HUMAN_Chan_2020")

print(ace2_dms)
```

The heatmap shows the DMS score at each position along the given protein
(x-axis) where a residue was mutated (alternate amino acid on y-axis). For this 
demonstration, we subset to the first 100 positions and grouped the amino
acids by their physiochemical properties (DE,KRH,NQ,ST,PGAVIL,MC,FYW). See
[here][physiochem] for more information. As a note, not all positions along the 
protein sequence may be subjected to mutation for every DMS assay. This results 
from the specific research objectives, prioritization choices of the 
investigators, or technical constraints inherent to the experimental design. 

A low DMS score indicates low fitness, while a higher DMS score indicates high 
fitness. We can think of higher DMS scores as being more benign, while lower DMS
score indicates more pathogenic regions.

Based on the "ACE2_HUMAN_Chan_2020" assay, we can see that at positions 90 and 
92, fitness remained high despite across amino acid changes; possibly 
suggestive of a benign region of the protein. However, several mutations at 
position 48 resulted in low fitness. This could represent an important region 
for protein function where any perturbation would likely be deleterious.

Let's plot another assay, specifying a region and invoking the `ComplexHeatmap`
row clustering under the hood. For more details about this clustering method or
to view more function parameters, read the function documentation with 
`?plot_dms_heatmap()`.

```{r SHOC2 heatmap, fig.wide = TRUE}
plot_dms_heatmap(assay_name = "SHOC2_HUMAN_Kwon_2022", 
    start_pos = 10,
    end_pos = 60,
    cluster_rows = TRUE)
```
In this region of the SHOC2_HUMAN protein, mutating to a K seemed to have the 
most benign affect across all positions.


# Visualize model scores along a protein

ProteinGymR Bioc 3.21 provides functionality to generate heatmaps of zero-shot
mode scores for over 70 variant effect prediction models with the function
`plot_zeroshot_heatmap()`. The only required arguments for this function are 
the assay name to plot (same as for the DMS heatmap), and a model to plot. For
a complete list of models, run `available_models()`. If `model_data` is not
provided, the default model scores from ProteinGym will be loaded in from
`ProteinGymR::zeroshot_substitutions()`.

```{r}
ace2_model <- plot_zeroshot_heatmap(
    assay_name = "ACE2_HUMAN_Chan_2020", 
    model = "GEMME")

print(ace2_model)
```

As with the DMS scores, we are plotting the GEMME zero-shot scores for each 
position available in the assay "ACE2_HUMAN_Chan_2020". Note that the model 
score here are mostly negative and thus, the color scale is red. Model scores
are not rescaled or normalized across the 78 models. For more information on
model scores and how to interpret them, consult the original ProteinGym 
[publication][pg_publication].

It was useful to visualize the DMS and model scores side by side for a given 
assay to compare the experimental DMS scores and predicted zero-shot scores 
outputted from the model. This is easily done in the given way:

```{r}
ComplexHeatmap::draw(ace2_dms %v% ace2_model)
```

`%v%` stacks the heatmaps in one column, while `+` will display them in two
columns, side by side. This can also be done with any output of class
ComplexHeatmap::Heatmap().


# 3D protein structure

The function requires a file path to a PBD file contain protein structure 
coordinates and a DMS assay to aggregate DMS scores that will be projected 
onto the 3D structure.

This vignette demonstrates how to explore and visualize DMS scores on a 3D 
protein structure using the package r3dmol under the hood. The function requires
a file path to a PBD file contain protein structure coordinates and a DMS assay
to aggregate DMS scores that will be projected onto the 3D structure.



By default, if no `dms_data` argument is provided, the DMS subsitutions from 
`ProteinGymR::dms_substitutions()` are loaded in. Alternatively, a user can 
specify their own dms list to project. The user may also specify what
aggregation method to use for calucating the statistic at each residue position.
By default, the mean DMS score is calculated for each position. See the function
documentation for details: `?plot_structure()`

First, let's plot using all default settings. The only required arguments are
the `assay_name` and the `pdb_file` path.

Importantly, the function normalizes the aggregated DMS scores so that a score
of zero will always be represented as white and the range of scores will be
scaled relatively to -1 and 1 in that assay, with 1 being the maximum 
aggregated DMS score within that assay region.

```{r}
plot_structure(assay_name = "ACE2_HUMAN_Chan_2020", 
    pdb_file = "~/ProteinGym_data/ProteinGym_v1.1/ProteinGym_AF2_structures/ACE2_HUMAN.pdb")
```

<br>

In this example, we are plotting the 3D structure of the ACE2_HUMAN protein and
overlaying the mean DMS score across all mutants in a given position. Chan et 
al. 2020 who generated the DMS assay data only experimentally assessed a subset
of the entire ACE2_HUMAN protein. By default the function only displayed the 
regions where there is information available in the assay. Red colors represent
more pathogenic (lower DMS scores) and blue colors show more benign positions
(higher DMS scores). Regions that appear white indicate closer to no change 
before and after the DMS perturbation.

If we want to visualize the complete protein structure, but simple grey out the
regions we do not have DMS assay information for, we can set the argument
`full_structure == TRUE`

```{r}
plot_structure(assay_name = "ACE2_HUMAN_Chan_2020", 
    pdb_file = "~/ProteinGym_data/ProteinGym_v1.1/ProteinGym_AF2_structures/ACE2_HUMAN.pdb",
    full_structure = TRUE)
```

<br>

Now we can more clearly see the entire protein structure for ACE2_HUMAN. 

Some assays extensively assessed nearly every position of the complete protein, 
for example: the C6KNH7_9INFA protein from Lee et al. 2018. Let's visualize 
this protein and set the aggregation method to view the minimum DMS score across
all mutants at each position by setting `aggregation_fun = min`.

```{r}
plot_structure(assay_name = "C6KNH7_9INFA_Lee_2018", 
    pdb_file = "~/ProteinGym_data/ProteinGym_v1.1/ProteinGym_AF2_structures/C6KNH7_9INFA.pdb",
    aggregate_fun = min)
```

<br>

For this protein, it looks like 

```{r}
dms_data <- dms_substitutions()
df <- dms_data[["C6KNH7_9INFA_Lee_2018"]]
hist(df$DMS_score)
```

We can see that for this DMS assay, most values of the minimum DMS score at each
position are negative and are thus represented as red.

<br>

Finally, the user can also zoom into a specific region of the protein by
defining residue position coordinates with `start_pos` and `end_pos`. 

Let's just visualize the mean DMS scores of the protein from position 10:50.

```{r}
plot_structure(assay_name = "ACE2_HUMAN_Chan_2020", 
    pdb_file = "~/ProteinGym_data/ProteinGym_v1.1/ProteinGym_AF2_structures/ACE2_HUMAN.pdb",
    start_pos = 10,
    end_pos = 50, full_structure = FALSE)
```


# Correlate DMS scores with model scores

By default, `dms_corr_plot()` allows the user to evaluate the correlation 
between experimental and model prediction scores. By default, it takes in a
protein UniProt ID and runs a Spearman correlation between the ProteinGym DMS 
assay scores and AlphaMissense predicted pathogenicity scores. It returns a 
ggplot for visualization.

```{r}
dms_corr_plot(uniprotId = "Q9NV35")
```

## Session Info
```{r}
sessionInfo()
```

