---
title: "Visualization and exploration"
author: 
    - name: Tram Nguyen
      affiliation: Department of Biomedical Informatics, Harvard Medical School
      email: Tram_Nguyen@hms.harvard.edu
    - name: Pascal Notin
      affiliation: Department of Systems Biology, Harvard Medical School
    - name: Aaron W Kollasch
      affiliation: Department of Systems Biology, Harvard Medical School
    - name: Debora Marks
      affiliation: Department of Systems Biology, Harvard Medical School
    - name: Ludwig Geistlinger
      affiliation: Department of Biomedical Informatics, Harvard Medical School
package: ProteinGymR
output:
    BiocStyle::html_document:
      self_contained: yes 
      toc: true
      toc_float: true
      toc_depth: 2
      code_folding: show
date: "April 23, 2025"
vignette: >
    %\VignetteIndexEntry{Visualization and exploration"}
    %\VignetteEncoding{UTF-8}
    %\VignetteEngine{knitr::rmarkdown}
editor_options: 
    markdown: 
      wrap: 80
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, nobreak = TRUE)
```

[CH]: https://bioconductor.org/packages/release/bioc/html/ComplexHeatmap.html
[r3d]: https://cran.r-project.org/web/packages/r3dmol/vignettes/r3dmol.html
[pg_publication]: https://proceedings.neurips.cc/paper_files/paper/2023/hash/cac723e5ff29f65e3fcbb0739ae91bee-Abstract-Datasets_and_Benchmarks.html

# Setup

```{r, message=FALSE, warning=FALSE, echo = FALSE}
library(ProteinGymR)
library(ComplexHeatmap)
library(stringr)
library(dplyr)
library(ggplot2)
library(ggExtra)
```


# Introduction 

This vignette demonstrates how to explore and visualize DMS and model scores 
from the ProteinGym database v1.2. Specifically, it walks through the 
functionality to generate heatmaps of all pairwise DMS substitution mutants and 
projects these scores onto 3D protein structures. `ProteinGymR` uses 
functionality from the Bioconductor package [ComplexHeatmap][CH] and 
[r3dmol][r3d] from CRAN under the hood to generate the heatmaps and 3D protein 
structures, respectively. 


# Visualize DMS scores along a protein

Explore the "ACE2_HUMAN" assay from Chan et al. 2020 and create a heatmap of 
the DMS scores with `plot_dms_heatmap()`. If the argument `dms_data` is not 
specified, the default will load the most recent DMS substitution data from 
ProteinGym with `ProteinGymR::dms_substitutions()`. This function only requires
a specific assay name. To obtain all assay names, run: 
`names(dms_substitutions())`. By default, the function also plots the full range
of positions where DMS data is available for this assay. To plot a specific 
region of interest, use the arguments `start_pos()` and `end_pos()` which takes
in an integer for the first and last residue position to plot in the protein.

```{r ACE2 default heatmap, fig.wide = TRUE}
ace2_dms <- plot_dms_heatmap(
    assay_name = "ACE2_HUMAN_Chan_2020",
    start_pos = 1, 
    end_pos = 100)

print(ace2_dms)
```

The heatmap shows the DMS score at each position along the given protein
(x-axis) where a residue was mutated (alternate amino acid on displayed on the 
y-axis and the reference allele at the position is shown on top). For this 
demonstration, we subset to the first 1-100 positions and grouped the amino
acids by their physiochemical properties (DE,KRH,NQ,ST,PGAVIL,MC,FYW). See
[here][physiochem] for more information. As a note, not all positions along the 
protein sequence may be subjected to mutation for every DMS assay. This results 
from the specific research objectives, prioritization choices of the 
investigators, or technical constraints inherent to the experimental design. 

A low DMS score indicates low fitness, while a higher DMS score indicates high 
fitness. We can think of higher DMS scores as being more benign, while lower DMS
score indicates more pathogenic regions.

Based on the "ACE2_HUMAN_Chan_2020" assay, we can see that at positions 90 and 
92, fitness remained high despite across amino acid changes; possibly 
suggestive of a benign region of the protein. However, several mutations at 
position 48 resulted in low fitness. This could represent an important region 
for protein function where any perturbation would likely be deleterious.

Let's plot another assay, specifying a region and invoking the `ComplexHeatmap`
row clustering under the hood. For more details about this clustering method or
to view more function parameters, read the function documentation with 
`?plot_dms_heatmap()`.

```{r SHOC2 heatmap, fig.wide = TRUE}
plot_dms_heatmap(assay_name = "SHOC2_HUMAN_Kwon_2022", 
    start_pos = 10,
    end_pos = 60,
    cluster_rows = TRUE)
```
In this region of the SHOC2_HUMAN protein, mutating to a K (y-axis) seemed to 
have the most benign affect across all mutations.


# Visualize model scores along a protein

ProteinGymR Bioc 3.21 provides functionality to generate heatmaps of zero-shot
mode scores for 79 variant effect prediction models and 11 semi-supervised 
models with the function `plot_zeroshot_heatmap()`. The required arguments 
for this function are the assay name to plot (same as for the DMS heatmap), 
and a model to plot. For a complete list of models, run `available_models()` for
zero-shot models, and `supervised_available_models()` for the 11 semi-supervised
models. If `model_data` is not provided, the default model scores from 
ProteinGym will be loaded in from `ProteinGymR::zeroshot_substitutions()`.

```{r, fig.wide = TRUE}
ace2_model <- plot_zeroshot_heatmap(
    assay_name = "ACE2_HUMAN_Chan_2020", 
    model = "GEMME",
    start_pos = 1,
    end_pos = 100)

print(ace2_model)
```

As with the DMS scores, we are plotting the GEMME zero-shot scores for each 
position available in the assay "ACE2_HUMAN_Chan_2020". Note that the model 
scores here are mostly negative and thus, the color scale is red. Model scores
are not rescaled or normalized across the 79 models, and comparison across of 
raw model predictions should be cautioned in this context. For more information 
on model scores and how to interpret them, consult the original ProteinGym 
[publication][pg_publication].

It can be useful to visualize the DMS and model scores side by side for a given 
assay to compare the experimental DMS scores and predicted zero-shot scores 
outputted from the model. This is easily done with the following:

```{r, fig.wide = TRUE}
ComplexHeatmap::draw(ace2_dms %v% ace2_model)
```

`%v%` stacks the heatmaps in one column, while `+` will display them in two
columns, side by side. This can also be done with any output of class
ComplexHeatmap::Heatmap().

<br>

# 3D protein structure

This vignette demonstrates how to explore and visualize DMS or model scores on 
a 3D protein structure using the package r3dmol under the hood. The function 
requires a file path to a PBD file containing protein structure coordinates and 
a DMS or model assay to aggregate DMS scores that will be projected onto the 3D 
structure.

By default, if no `dms_data` argument is provided, the DMS substitutions from 
`ProteinGymR::dms_substitutions()` are loaded in, or if viewing model scores, 
model scores from an assay loaded with `ProteinGymR::model_substitutions()` will
be used as default. Alternatively, a user can specify their own DMS list to 
project. The user may also specify what aggregation method to use for 
calculating the summary statistic at each residue position. By default, 
the mean DMS score/model prediction score is calculated for each position. 
See the function documentation for details: `?plot_structure()`

First, let's use all the default settings. The only required arguments are
the `assay_name` and the `pdb_file` file path.

Importantly, the function normalizes the aggregated DMS scores so that a score
of zero will always be represented as white and the range of scores will be
scaled relatively to -1 and 1 in that assay, with 1 being the maximum 
aggregated DMS score within that assay region.

```{r, fig.wide = TRUE}
plot_structure(assay_name = "ACE2_HUMAN_Chan_2020")
```

<br>

In this example, we are plotting the 3D structure of the ACE2_HUMAN protein and
overlaying the mean DMS score across all mutants in a given position. Chan et 
al. 2020 who generated the DMS assay data only experimentally assessed a subset
of the entire ACE2_HUMAN protein. By default the function only displayed the 
regions where there is information available in the assay. Red colors represent
more pathogenic (lower DMS scores) and blue colors show more benign positions
(higher DMS scores). Regions that appear white indicate closer to no change 
before and after the DMS perturbation. By default, regions of the protein 
without experiment information have the "ball and stick" representation.

If we want to visualize the complete protein structure, but simply grey out the
regions we do not have DMS assay information for rather than using the 
"ball and stick" representation, we can set the argument 
`full_structure == TRUE`.

<br>

```{r, fig.wide = TRUE}
plot_structure(assay_name = "ACE2_HUMAN_Chan_2020", 
    full_structure = TRUE)
```

<br>

Now we can more clearly see the entire protein structure for ACE2_HUMAN in the 
ribbon representation. 

Some assays extensively assessed nearly every position of the complete protein, 
for example: the C6KNH7_9INFA protein from Lee et al. 2018. Let's visualize 
this protein and set the aggregation method to view the minimum DMS score across
all mutants at each position by setting `aggregation_fun = min`.

```{r, fig.wide = TRUE}
plot_structure(assay_name = "C6KNH7_9INFA_Lee_2018", 
    aggregate_fun = min)
```

<br>

As we might expect, the minimum DMS value (more pathogenic) is almost always a 
negative number across all positions of this protein. Therefore, there seems to 
be at least one amino acid mutation that could severely distrupt the fitness at
any position of this protein.

```{r, fig.wide = TRUE}
dms_data <- dms_substitutions()
df <- dms_data[["C6KNH7_9INFA_Lee_2018"]]

ggplot(df, aes(DMS_score)) +
    geom_histogram() + 
    theme_minimal()
```

<br> 
We can see that for this DMS assay, most values of the minimum DMS score at each
position are negative and are thus represented as red.

<br>

Finally, the user can also zoom into a specific region of the protein by
defining residue position coordinates with `start_pos` and `end_pos`. 

Let's just visualize the mean DMS scores of the protein from position 10:50.

<br>

```{r, fig.wide = TRUE}
plot_structure(assay_name = "ACE2_HUMAN_Chan_2020", 
    start_pos = 10,
    end_pos = 50, 
    full_structure = FALSE)
```


Finally, it is possible to use the same color scheme as the popEVE 
mutation [portal](https://pop.evemodel.org/). We can do this for any of 
the heatmaps or protein structure plots.

```{r, fig.wide = TRUE}
plot_structure(assay_name = "ACE2_HUMAN_Chan_2020",
    color_scheme = "EVE")
```


# Correlate DMS scores with model scores

By default, `dms_corr_plot()` allows the user to evaluate the correlation 
between experimental and model prediction scores. By default, it takes in a
protein UniProt ID and runs a Spearman correlation between the ProteinGym DMS 
assay scores and AlphaMissense predicted pathogenicity scores. It returns a 
ggplot for visualization.

<br>

```{r, warning = FALSE, fig.wide = TRUE}
dms_corr_plot(uniprotId = "Q9NV35")
```

<br>

Be default, the `dms_corr_plot()` function gathers any of the 217 DMS assays of 
the chosen UniProt ID and correlates the average DMS score across relevant
assays and the AlphaMissense model predictions.

```{r}
dms_corr_plot(uniprotId = "Q9NV35")
```

Be default, the `dms_corr_plot()` function gathers any of the 217 DMS assays of 
the chosen UniProt ID and correlates the average DMS scoree across relevant
assays and the AlphaMissense model predictions.

<br>

Although the default uses the AlphaMissense scores, it is also possible to 
correlate DMS experimental scores with predictions from any of the 79 zero-shot 
or 11 supervised models. Below is an example of the workflow to accomplish this
for a single assay "A0A140D2T1_ZIKV_Sourisseau_2019".

Load in both the zero-shot and DMS substitution scores for the 217 assays.

```{r}
## Load model scores
model_data <- zeroshot_substitutions()

## Load DMS scores
dms_data <- dms_substitutions()
```

Specify the chosen assay and model to explore. Again, to view all assay names to 
choose from, run `names()` on the list loaded in by `dms_substitutions()`.

```{r}
## Specify the chosen assay and model to explore.
assay_name <- "A0A140D2T1_ZIKV_Sourisseau_2019"
model <- "GEMME"

## Filter selected assay
model_df <- as.data.frame(model_data[[assay_name]])
dms_df <- as.data.frame(dms_data[[assay_name]])

## Filter out multiple aa sites
model_df <- model_df |>
    filter(!str_detect(.data$mutant, ":"))

dms_df <- dms_df |>  
    filter(!str_detect(.data$mutant, ":"))

## Select chosen model
model_df <- model_df |>
    dplyr::select(
        mutant,
        all_of(model)
    )

model_df <- model_df |>
    dplyr::select(all_of(model), mutant)
```

Now we wrangle and merge the scores into a data.frame that we will use for 
plotting.

```{r, fig.wide = TRUE}
## Wrangle the data
dms_df <- dms_df |>
   dplyr::select(DMS_score, mutant)

merged_table <- 
        left_join(
            model_df, dms_df, 
            by = c("mutant"),
            relationship = "many-to-many"
        ) |>
        na.omit() |> 
        select(mutant, DMS_score, model = all_of(model))

## Corr function
pg_correlate <- 
    function(merged_table)
{
    cor_results <- 
        cor.test(
            merged_table$DMS_score, merged_table$model, 
            method=c("spearman"), 
            exact = FALSE
        )
    
    cor_results
}

cor_results <- pg_correlate(merged_table)

## Correlation density plot
pg_density_plot <- 
    merged_table |> 
    ggplot(
        aes(x = .data$DMS_score, y = .data$model)
    ) +
    geom_bin2d(bins = 60) +
    geom_point(alpha = 0) +
    scale_fill_continuous(type = "viridis") +
    xlab("DMS score") +
    ylab(paste0(model, "\nzero shot score")) +
    theme_classic() +
    theme(
        axis.text.x = element_text(size = 16),
        axis.text.y = element_text(size = 16),
        axis.title.y = element_text(size = 16, vjust = 2),
        axis.title.x = element_text(size = 16, vjust = 0),
        legend.title = element_text(size = 16),
        legend.text = element_text(size = 16)
    )

# Add marginal density plots
pg_density_plot <- ggMarginal(
        pg_density_plot,
        type = "densigram", # Can also use "histogram"
        fill = "#B0C4DE", 
        color = "black"  # Change color as needed
    )


pg_density_plot

print(paste0("r = ", format(round(cor_results$estimate, 2)), 
                "; Pval = ", cor_results$p.value))
```

<br>

We can also explore the correlation between two different models.

```{r, fig.wide = TRUE}
## Select assay and models to compare
assay_name <- "A0A140D2T1_ZIKV_Sourisseau_2019"
model1 <- "EVE_ensemble"
model2 <- "S3F_MSA"

## Filter selected assay
model_df <- model_data[[assay_name]]

## Filter out multiple aa sites
model_df <- model_df |>  
    filter(!grepl(":", .data$mutant))

## Select chosen model
model_df <- model_df |>
    dplyr::select(
        mutant,
        all_of(c(model1 = model1, model2 = model2))
    )

## Corr function
pg_correlate <- 
    function(model_df)
{
    cor_results <- 
        cor.test(
            model_df$model1, model_df$model2, 
            method=c("spearman"), 
            exact = FALSE
        )
    
    cor_results
}

cor_results <- pg_correlate(model_df = model_df)

## Correlation density plot
pg_density_plot <- 
    model_df |> 
    ggplot(
        aes(x = .data$model1, y = .data$model2)
    ) +
    geom_bin2d(bins = 60) +
    geom_point(alpha = 0) +
    scale_fill_continuous(type = "viridis") +
    xlab(paste0(model1, "\nzero shot score")) +
    ylab(paste0(model2, "\nzero shot score")) +
    theme_classic() +
    theme(
        axis.text.x = element_text(size = 16),
        axis.text.y = element_text(size = 16),
        axis.title.y = element_text(size = 16, vjust = 2),
        axis.title.x = element_text(size = 16, vjust = 0),
        legend.title = element_text(size = 16),
        legend.text = element_text(size = 16)
    )

## Add marginal density plots
pg_density_plot <- ggMarginal(
        pg_density_plot,
        type = "densigram", # Can also use "histogram"
        fill = "#B0C4DE", 
        color = "black"  # Change color as needed
        )
    
pg_density_plot

print(paste0("r = ", format(round(cor_results$estimate, 2)), 
                "; Pval = ", cor_results$p.value))
```
<br>

Between the EVE_ensemble and S3F_MSA, there seems to be good correlation between
the model predictions for the variants in the assay 
"A0A140D2T1_ZIKV_Sourisseau_2019".

## Session Info
```{r}
sessionInfo()
```