---
title: "make_ProteinGym_DMS_substitutions"
author: "Tram Nguyen"
date: "2024-08-12"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r, warning=FALSE}
suppressMessages(library(dplyr))
suppressMessages(library(readr))
suppressMessages(library(UniProt.ws))
suppressMessages(library(org.Hs.eg.db))
```

# Download data

All datasets were downloaded from
[ProteinGym.org](%22ProteinGym.org%22). Specifically, the substitution
data were accessed from
<https://marks.hms.harvard.edu/tranception/ProteinGym_substitutions.zip>
(2024-07-22).


# ProteinGym information

ProteinGym is a collection of benchmarks aiming at comparing the ability
of models to predict the effects of protein mutations, and were curated by 
Notin et al. 
[2023](https://papers.nips.cc/paper_files/paper/2023/hash/cac723e5ff29f65e3fcbb0739ae91bee-Abstract-Datasets_and_Benchmarks.html). 
One component of ProteinGym is the Deep Mutational Scanning (DMS) assays
that can serve as ground truth for assessing variant predictions.

The substitution benchmark consists of more than 2.5M variants from 217
assays.
We provide these assays in a list object, with the exception of the 
"ANCSZ_Hobbs_2022" assay, which comprised a manual protein reconstruction by
the original authors and thus had no matching UniProt accession. Therefore,
this list contains a total of 216 elements, each corresponding to a
data.frame for a single DMS assay. Each data.frame contains the following
columns:

**UniProt_id (str):** UniProt accession identifier for the protein.

**DMS_id (str):** DMS assay identifier, describing the protein and original
authors generating the experiment data. This field matches the names of the 
original .csv files downloaded from ProteinGym.org for each corresponding assay.

**mutant (str):** describes the set of substitutions to apply on the
reference sequence to obtain the mutated sequence (eg., A1P:D2N implies
the amino acid 'A' at position 1 should be replaced by 'P', and 'D' at
position 2 should be replaced by 'N').

**mutated_sequence (str):** represents the full amino acid sequence for
the mutated protein.

**DMS_score (float):** corresponds to the experimental measurement in
the DMS assay. Higher DMS_score values are associated with higher fitness of
the mutated protein.

**DMS_score_bin (int):** indicates whether the DMS_score is above the
fitness cutoff defined by ProteinGym (1 is fit, 0 is not fit).

# Load in all ProteinGym assays

```{r, warning=FALSE}
# Downloaded manually from ProteinGym website 
# https://proteingym.org/download
# Data stored locally
MAVE.dir <- "/home/rstudio/AlphaMissense_data/DMS_ProteinGym_substitutions/"

ProGym_files <- list.files(path=MAVE.dir, full.names=F)

# Grab MAVE study names. These should match AlphaMissense.
ProGym_names <- sub(".csv", "", ProGym_files) 

# Load in all the files as tables in a list
ProGym_tables <- suppressMessages(lapply(paste0(MAVE.dir, ProGym_files), 
                                         read_csv, show_col_types = FALSE))

# Assign respective MAVE study names to each list element
names(ProGym_tables) <- ProGym_names

# Combine all of the dataframes together with each df name as a new column:
# `DMS_id` is the protein entry name with ProteinGym experiment names
add_name_column <- function(df, name) {
  df %>% mutate(DMS_id = name)
}

# Apply the function to each data frame in the list
ProGym_tables <- lapply(names(ProGym_tables), function(name) {
  add_name_column(ProGym_tables[[name]], name)
})

# Reassign the names to the list
names(ProGym_tables) <- ProGym_names
```

# Convert ProteinGym's Entry Names to UniProt Accession

```{r}
# Grab the protein Entry Name from ProteinGym assays by extracting only the 
entryNames <- strsplit(names(ProGym_tables), split = "_", fixed = TRUE)

# Function to grab SwissProt ID from DMS names (first and second elements)
entryNames <- sapply(entryNames, function(input_string) {
  first_two_elements <- paste(input_string[1], input_string[2], sep = "_")
  return(first_two_elements)
})

# Convert entryNames to UniProt Accession ID
ws <- UniProt.ws()
out <- select(ws, entryNames, columns = "UniProtKB", keytype = "UniProtKB")
accessions <- out$Entry

# Manually checked results of mapping: Merge results with original PG studies
pg_names <- as.data.frame(entryNames)
cdf <- left_join(pg_names, out, by = c("entryNames" = "From"), keep = TRUE)
colnames(cdf) <- c("pg_names", "entry", "accessions")
```

# Two IDs not mapped: ANCSZ_Hobbs, PSAE_SYNP2

```{r}
# Manual curation of PSAE_SYNP2: to UniProt ID "P31969"
cdf$accessions[which(cdf$pg_names == "PSAE_SYNP2")] <- "P31969"
cdf$entry[which(cdf$pg_names == "PSAE_SYNP2")] <- "PSAE_SYNP2"
```

The "ANCSZ_Hobbs" assay is a DMS of an ancestral reconstruction of the
Syk-family kinases, AncSZ, that [Hobbs et al.
2022](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC9601881/) manually
curated. It is an ancestral sequence reconstruction to generate a spleen
tyrosine kinase (Syk), and therefore does not have an established
UniProt accession ID. For this protein, we will remove this assay from
our data, resulting in 216 total DMS assays for this resource.

```{r}
# Remove it from the accession list
cdf <- na.omit(cdf)

# Remove ANCSZ_Hobbs
ProGym_tables[["ANCSZ_Hobbs_2022"]] <- NULL

# update the PG study names without Hobbs -- Used later
ProGym_names <- names(ProGym_tables)
```

# Add a column in the ProteinGym data with UniProt Accession

```{r}
# Assign respective UniProt names to each list element - to do next function
names(ProGym_tables) <- cdf$accessions

# For each assay, mutate a new UniProt ID column
add_name_column <- function(df, name) {
  df %>% mutate(UniProt_id = name)
}

# Apply the function to each data frame in the list
ProGym_tables <- lapply(names(ProGym_tables), function(name) {
  add_name_column(ProGym_tables[[name]], name)
})

# Reassign the DMS study names to the list elements
names(ProGym_tables) <- ProGym_names
```

```{r}
# Reorder the columns
cols <- c("UniProt_id", "DMS_id", "mutant", "mutated_sequence", 
          "DMS_score", "DMS_score_bin")

ProGym_tables <- lapply(ProGym_tables, function(df) {
   df %>% dplyr::select(all_of(cols))
})

# Make "DMS_score_bin" factor
ProGym_tables <- lapply(ProGym_tables, function(df) {
  df %>%
    mutate(DMS_score_bin = factor(DMS_score_bin))
})
```

Returns a list of 216 unique DMS assays, each with 6 columns:
"UniProt_id","DMS_id","mutant","mutated_sequence","DMS_score","DMS_score_bin".


# Write ProteinGym DMS list to RDS object

```{r, eval=FALSE}
saveRDS(ProGym_tables, "ProteinGym/ProGym216_DMS_subs_v1.rds"))
```

# SessionInfo

```{r}
sessionInfo()
```
