---
title: "make_ProteinGym_DMS_substitutions"
author: "Tram Nguyen"
date: "2024-10-01"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r, warning=FALSE}
suppressMessages(library(dplyr))
suppressMessages(library(readr))
suppressMessages(library(UniProt.ws))
suppressMessages(library(org.Hs.eg.db))
```

# Download data

All datasets were downloaded from
[ProteinGym.org](%22ProteinGym.org%22). Specifically, the substitution
data were accessed from
<https://marks.hms.harvard.edu/tranception/ProteinGym_substitutions.zip>
(2024-07-22).

The reference file for these DMS assays (metadata) were downloaded at
<https://marks.hms.harvard.edu/proteingym/DMS_substitutions.csv>
(2024-10-01)

# ProteinGym information

ProteinGym is a collection of benchmarks aiming at comparing the ability
of models to predict the effects of protein mutations, and were curated
by Notin et al. 
[2023](https://papers.nips.cc/paper_files/paper/2023/hash/cac723e5ff29f65e3fcbb0739ae91bee-Abstract-Datasets_and_Benchmarks.html). 
One component of ProteinGym is the Deep Mutational Scanning (DMS) assays
that can serve as ground truth for assessing variant predictions.

The substitution benchmark consists of more than 2.5M variants from 217
assays. We provide these assays in a `list` object with each element
corresponding to a `data.frame` for each DMS assay. 

# Load in all ProteinGym assays

```{r, warning=FALSE}
# Downloaded manually from ProteinGym website 
# https://proteingym.org/download
# Data stored locally
MAVE_dir <- "/home/rstudio/ProteinGym_data/DMS_substitutions/DMS_ProteinGym_substitutions/"

ProGym_files <- list.files(path=MAVE_dir, full.names=F)

# Grab MAVE study names. These should match AlphaMissense.
ProGym_names <- sub(".csv", "", ProGym_files) 

# Load in all the files as tables in a list
ProGym_tables <- suppressMessages(lapply(paste0(MAVE_dir, ProGym_files),
                                         read_csv, show_col_types = FALSE))

# Assign respective MAVE study names to each list element
names(ProGym_tables) <- ProGym_names

# Combine all of the dataframes together with each df name as a new column:
# `DMS_id` is the protein entry name with ProteinGym experiment names
ProGym_tables <- 
    purrr::map2(
        ProGym_tables, 
        ProGym_names, ~ {.x |>
            mutate(DMS_id = .y)
        }
    )
```

# Convert ProteinGym's Entry Names to UniProt Accession

```{r}
# Grab the protein Entry Name from ProteinGym assays by extracting only the
entryNames <- strsplit(names(ProGym_tables), split = "_", fixed = TRUE)

# Function to grab SwissProt ID from DMS names (first and second elements)
entryNames <- sapply(entryNames, function(input_string) {
  first_two_elements <- paste(input_string[1], input_string[2], sep = "_")
  return(first_two_elements)
})

# Convert entryNames to UniProt Accession ID
ws <- UniProt.ws()
out <- select(ws, entryNames, columns = "UniProtKB", keytype = "UniProtKB")
accessions <- out$Entry

# Manually checked results of mapping: Merge results with original PG studies
pg_names <- as.data.frame(entryNames)
cdf <- left_join(pg_names, out, by = c("entryNames" = "From"), keep = TRUE)
colnames(cdf) <- c("pg_names", "entry", "accessions")
```

# Two IDs not mapped: ANCSZ_Hobbs, PSAE_SYNP2
# Manually map

```{r}
# Manual curation of PSAE_SYNP2: to UniProt ID "P31969"
cdf$accessions[which(cdf$pg_names == "PSAE_SYNP2")] <- "P31969"
cdf$entry[which(cdf$pg_names == "PSAE_SYNP2")] <- "PSAE_SYNP2"

# Manually curation of ANCZ -- leave UniProt accession as NA
cdf$entry[which(cdf$pg_names == "ANCSZ_Hobbs")] <- NA
```

# Add a column in the ProteinGym data with UniProt Accession

```{r}
ProGym_names <- names(ProGym_tables)

# For each assay, mutate a new UniProt ID column
uniprot_vec <- cdf$accessions

ProGym_tables <- 
    purrr::map2(ProGym_tables, 
        uniprot_vec, ~ { .x |>
            mutate(UniProt_id = .y)
        }
    )

# Reorder the columns
cols <- c("UniProt_id", "DMS_id", "mutant", "mutated_sequence", 
          "DMS_score", "DMS_score_bin")

ProGym_tables <- lapply(ProGym_tables, function(df) {
   df |>  dplyr::select(all_of(cols))
})

# Make "DMS_score_bin" factor
ProGym_tables <- lapply(ProGym_tables, function(df) {
    df |>
    mutate(DMS_score_bin = factor(DMS_score_bin))
})
```

Changed tibble to data.frames.
```{r}
ProGym_tables <- lapply(ProGym_tables, as.data.frame)
```

Returns a list of 216 unique DMS assays, each with 6 columns:
"UniProt_id","DMS_id","mutant","mutated_sequence","DMS_score","DMS_score_bin".

**UniProt_id (str):** UniProt accession identifier for the protein.

**DMS_id (str):** DMS assay identifier, describing the protein and original
authors generating the experiment data. This field matches the names of the 
original .csv files downloaded from ProteinGym.org for each corresponding assay.

**mutant (str):** describes the set of substitutions to apply on the
reference sequence to obtain the mutated sequence (eg., A1P:D2N implies
the amino acid 'A' at position 1 should be replaced by 'P', and 'D' at
position 2 should be replaced by 'N').

**mutated_sequence (str):** represents the full amino acid sequence for
the mutated protein.

**DMS_score (float):** corresponds to the experimental measurement in
the DMS assay. Higher DMS_score values are associated with higher fitness of
the mutated protein.

**DMS_score_bin (int):** indicates whether the DMS_score is above the
fitness cutoff defined by ProteinGym (1 is fit, 0 is not fit).


# Write ProteinGym DMS list to RDS object

```{r, eval=FALSE}
saveRDS(ProGym_tables, 
    "/home/rstudio/ProteinGym_data/EH_data/v1.1/ProGym217_DMS_subs_v1.rds")
```


# Load and write metadata table
```{r, eval=FALSE}
ref_file <- read_csv(
    "/home/rstudio/ProteinGym_data/DMS_substitutions/DMS_substitutions.csv")

ref_file$taxon <- as.factor(ref_file$taxon)
ref_file$source_organism <- as.factor(ref_file$source_organism)
ref_file$DMS_binarization_method <- as.factor(ref_file$DMS_binarization_method)
ref_file$selection_type <- as.factor(ref_file$selection_type)
ref_file$selection_assay <- as.factor(ref_file$selection_assay)
ref_file$raw_DMS_phenotype_name <- as.factor(ref_file$raw_DMS_phenotype_name)
ref_file$raw_DMS_directionality <- as.factor(ref_file$raw_DMS_directionality)
ref_file$raw_DMS_mutant_column <- as.factor(ref_file$raw_DMS_mutant_column)
ref_file$ProteinGym_version <- as.factor(ref_file$ProteinGym_version)
ref_file$coarse_selection_type <- as.factor(ref_file$coarse_selection_type)

ref_file <- as.data.frame(ref_file)

saveRDS(ref_file, 
    "/home/rstudio/ProteinGym_data/EH_data/v1.1/ref_file_217_DMS_subs_v1.rds")
```

# SessionInfo

```{r}
sessionInfo()
```
