---
title: "make_zero_shot_substitutions_scores"
author: "Tram Nguyen"
date: "2025-03-26"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r, warning=FALSE}
suppressMessages(library(dplyr))
suppressMessages(library(readr))
suppressMessages(library(UniProt.ws))
suppressMessages(library(org.Hs.eg.db))
```

[Zenodo_v1.2]: https://zenodo.org/records/14997691
[Notin_pub]: https://papers.nips.cc/paper_files/paper/2023/hash/cac723e5ff29f65e3fcbb0739ae91bee-Abstract-Datasets_and_Benchmarks.html


# Download data

Zero shot scores for all DMS substitution in 217 assays were accessed from the [ProteinGym Zenodo][Zenodo_v1.2] repository (v1.2).
(accessed 2025-03-13).

```{r, warning=FALSE}
# Data stored locally
file_dir <- "/home/rstudio/ProteinGym_data/ProteinGym_v1.2/zero_shot_substitutions_scores/"

ProGym_files <- list.files(path=file_dir, full.names=FALSE)

# Grab assay names
ProGym_names <- sub(".csv", "", ProGym_files) 

# Load in all the files as tables in a list
ProGym_tables <- suppressMessages(lapply(paste0(file_dir, ProGym_files),
                                         read_csv, show_col_types = FALSE))

# Assign respective study names to each list element
names(ProGym_tables) <- ProGym_names

# Add column with `DMS_id` for each study respectively
ProGym_tables <- 
    purrr::map2(
        ProGym_tables, 
        ProGym_names, ~ {.x |>
            mutate(DMS_id = .y)
        }
    )
```

# Convert ProteinGym's Entry Names to UniProt Accession

```{r}
# Grab the protein Entry Name from ProteinGym assays by extracting only the
entryNames <- strsplit(names(ProGym_tables), split = "_", fixed = TRUE)

# Function to grab SwissProt ID from DMS names (first and second elements)
entryNames <- sapply(entryNames, function(input_string) {
  first_two_elements <- paste(input_string[1], input_string[2], sep = "_")
  return(first_two_elements)
})

# Convert entryNames to UniProt Accession ID
ws <- UniProt.ws()
out <- UniProt.ws::select(ws, entryNames, columns = "UniProtKB", 
    keytype = "UniProtKB")
accessions <- out$Entry

# Manually checked results of mapping: Merge results with original PG studies
pg_names <- as.data.frame(entryNames)
cdf <- left_join(pg_names, out, by = c("entryNames" = "From"), keep = TRUE)
colnames(cdf) <- c("pg_names", "entry", "accessions")
```

# Two IDs not mapped: ANCSZ_Hobbs
# Manually map

```{r}
# Manually curation of ANCZ -- leave UniProt accession as NA
cdf$entry[which(cdf$pg_names == "ANCSZ_Hobbs")] <- NA
```

# Add a column in the ProteinGym data with UniProt Accession

```{r}
ProGym_names <- names(ProGym_tables)

# For each assay, mutate a new UniProt ID column
uniprot_vec <- cdf$accessions

ProGym_tables <- 
    purrr::map2(ProGym_tables, 
        uniprot_vec, ~ { .x |>
            mutate(UniProt_id = .y)
        }
    )

# Reorder the columns
cols <- c("UniProt_id", "DMS_id", "mutant", "mutated_sequence", 
          "DMS_score", "DMS_score_bin")

ProGym_tables <- lapply(ProGym_tables, function(df) {
   df |>  dplyr::select(all_of(cols), everything())
})

# Make "DMS_score_bin" factor
ProGym_tables <- lapply(ProGym_tables, function(df) {
    df |>
    mutate(DMS_score_bin = factor(DMS_score_bin))
})

# Replace column hyphens with underscores
replace_hyphen <- function(df) {
    # Replace in column names
    names(df) <- gsub("-", "_", names(df))
    
    # Replace in all values (assuming all columns are character or factors)
    df[] <- lapply(df, function(col) {
    if (is.character(col)) {
      return(gsub("-", "_", col))
    } else if (is.factor(col)) {
      return(factor(gsub("-", "_", as.character(col))))
    } else {
      return(col)  # Keep other types unchanged
    }
    })
    
    return(df)
}

# Apply the function to each dataframe in the list
ProGym_tables <- lapply(ProGym_tables, replace_hyphen)
```

Changed tibble to data.frames.
```{r}
ProGym_tables <- lapply(ProGym_tables, as.data.frame)

# Remove extra bin_score column
ProGym_tables[["ANCSZ_Hobbs_2022"]][,7] <- NULL
```

Returns a list of multiple zero-shot model scores for each 217 DMS 
substitution assays.


# Write ProteinGym zero-shot list to RDS object

```{r, eval=FALSE}
saveRDS(ProGym_tables, 
    "/home/rstudio/ProteinGym_data/EH_data/v1.2/zeroshot_scores_v1.2.rds")
```

