---
title: "make_zeroshot_DMS_subs"
author: "Tram Nguyen"
date: "2024-08-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Zero-shot benchmarking information

In the zero-shot setting, experimental phenotypical measurements from
a given assay are predicted without having access to the labels at training
time. Model performance was evaluted across 5 metrics:

1. Spearman's rank correlation coefficient (primary metric)
2. Area Under the ROC Curve (AUC)
3. Matthews Correlation Coefficient (MCC) for bimodal DMS measurements
4. Normalized Discounted Cumulative Gains (NDCG) for identifying the most 
functional protein variants
5. Top K Recall (top 10% of DMS values)

To avoid placing too much weight on properties with many assays 
(e.g., thermostability), these metrics were first calculated within groups 
of assays that measure similar functions. The final value of the metric
is then the average of these averages, giving each functional group equal 
weight. The final values are referred to as the ‘corrected average’.

Due to the often non-linear relationship between protein function and 
organism fitness [Boucher et al., 2016](https://pubmed.ncbi.nlm.nih.gov/27010590/), 
the Spearman’s rank correlation coefficient is the most generally appropriate 
metric for model performance on experimental measurements. However, in 
situations where DMS measurements exhibit a bimodal profile, rank correlations 
may not be the optimal choice. Therefore, additional metrics are also provided, 
such as the Area Under the ROC Curve (AUC) and the Matthews Correlation 
Coefficient (MCC), which compare model scores with binarized experimental 
measurements. Furthermore, for certain goals (e.g., optimizing functional 
properties of designed proteins), it is more important that a model is able to 
correctly identify the most functional protein variants, rather than properly 
capture the overall distribution of all assayed variants. Thus, we also 
calculate the Normalized Discounted Cumulative Gains (NDCG), which up-weights a 
model if it gives its highest scores to sequences with the highest DMS value. 
Finally, we also calculate Top K Recall, where we select K to be the top 10% 
of DMS values.

For each metric, we provide a corresponding a .csv file. One loaded, each 
data.frame contains the corrected average values for 67 models (columns) across 
217 assays (rows) calculated on the DMS substitution provided in ProteinGym.

# Download and process data

The .csv files of benchmarking scores were downloaded from [the ProteinGym Github repo](https://github.com/OATML-Markslab/ProteinGym/tree/main/benchmarks/DMS_zero_shot/substitutions)
(accessed 2024-08-28).

Briefly, to calculate the DMS substitution benchmark metrics, the model scores 
are downloaded from [ProteinGym](https://marks.hms.harvard.edu/proteingym/zero_shot_substitutions_scores.zip).
The script to run the calculation is provided [here](https://github.com/OATML-Markslab/ProteinGym/tree/main/scripts/scoring_DMS_zero_shot/performance_substitutions.sh).

```{r, eval = FALSE}

url_prefix <- "https://github.com/OATML-Markslab/ProteinGym/blob/main/benchmarks/DMS_zero_shot/substitutions/"

## AUC
file_name <- "DMS_substitutions_AUC_DMS_level.csv"

download.file(
    url = paste0(
        url_prefix, "AUC/", file_name
        ), 
    destfile = paste0("/home/rstudio/ProteinGym_data/DMS_sub_benchmark_scores/", 
        file_name)
)

## MCC
file_name <- "DMS_substitutions_MCC_DMS_level.csv"

download.file(
    url = paste0(
        url_prefix, "MCC/", file_name
        ), 
    destfile = paste0("/home/rstudio/ProteinGym_data/DMS_sub_benchmark_scores/", 
        file_name)
)
    
## NDCG
file_name <- "DMS_substitutions_NDCG_DMS_level.csv"

download.file(
    url = paste0(
        url_prefix, "NDCG/", file_name
        ), 
    destfile = paste0("/home/rstudio/ProteinGym_data/DMS_sub_benchmark_scores/", 
        file_name)
)   

## Spearman
file_name <- "DMS_substitutions_Spearman_DMS_level.csv"

download.file(
    url = paste0(
        url_prefix, "Spearman/", file_name
        ), 
    destfile = paste0("/home/rstudio/ProteinGym_data/DMS_sub_benchmark_scores/", 
        file_name)
)

## Top_recall
file_name <- "DMS_substitutions_Top_recall_DMS_level.csv"

download.file(
    url = paste0(
        url_prefix, "Top_recall/", file_name
        ), 
    destfile = paste0("/home/rstudio/ProteinGym_data/DMS_sub_benchmark_scores/", 
        file_name)
)
```

# Reference

Notin, P., Kollasch, A., Ritter, D., van Niekerk, L., Paul, S., Spinner, H., 
Rollins, N., Shaw, A., Orenbuch, R., Weitzman, R., Frazer, J., Dias, M., 
Franceschi, D., Gal, Y., & Marks, D. (2023). ProteinGym: Large-Scale 
Benchmarks for Protein Fitness Prediction and Design. In A. Oh, T. Neumann, 
A. Globerson, K. Saenko, M. Hardt, & S. Levine (Eds.), Advances in Neural 
Information Processing Systems (Vol. 36, pp. 64331-64379). 
Curran Associates, Inc.

Boucher, J. I., Bolon, D. N., and Tawfik, D. S. Quantifying and understanding 
the fitness effects of protein mutations: Laboratory versus nature. 
Protein Science, 25(7):12191226, 2016.

# SessionInfo

```{r}
sessionInfo()
```
