---
title: "make_AlphaMissense_supplement"
author: "Tram Nguyen"
date: "2024-07-10"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(width = 80)
```

# Data information

Data from ProteinGym datasets was curated and processed by (19) to assess predictors on
predicting variant effects on protein fitness. The substitution benchmark consists of ~1.5M
variants from 87 MAVE experiments measured with 72 proteins. Data was accessed from
https://marks.hms.harvard.edu/tranception/ProteinGym_substitutions.zip (accessed 2023-01-13).
We provide the list of variants and our predictions as Supplementary Data S8.

Additional MAVE datasets
To generate an additional independent MAVE benchmark data set, we searched across MaveDB
(Esposito et al., 2019), MaveRegistry (Da Kuang et al., 2021), PubMed and BioRxiv. Studies
were selected for inclusion if they were (1) not originally in ProteinGym; (2) covered a human
protein; (3) measured a phenotype relevant to protein abundance/trafficking, protein-protein
interaction, enzymatic activity, or function; (4) had more than 100 variant measurements. MAVE
data was typically reported as a table available for download from the publication or data
repository. The exception was PPARG_HUMAN, for which we queried the webserver
accompanying the paper (https://miter.broadinstitute.org/) by searching through all possible
missense mutations at prevalence 1e-5. This was performed by changing the [ID] field in this
URL: https://miter.broadinstitute.org/mitergrade/?query=p.[ID]&prevalence=1.0e-5. Data
sources for these additional MAVE datasets are listed in Table S5.


# Download processed data

The original .csv table is available for download through the [AlphaMissense
Supplementary Data](https://www.science.org/doi/suppl/10.1126/science.adg7492/suppl_file/science.adg7492_data_s1_to_s9.zip).

Additional cell metadata, TSNE and PCA coordinates were provided 
courtesy of the authors.

# Generate `SingleCellExperiment` object

```{r}
library(dplyr)
library(tidyr)
library(SingleCellExperiment)

# Convert Seurat to SingleCellExperiment object
s_sce <- as.SingleCellExperiment(s2, assay="RNA")

# Format cell metadata
s_sce$orig.ident <- NULL
s_sce$ident <- NULL

cols <- c("animal", "batch", "animal_type", "cell_type")
colData(s_sce)[cols] <- lapply(colData(s_sce)[cols], factor)

# Add cell type colors matching paper
metadata(s_sce)$cell_colors<-s2@misc$colours

# Format reduced dim elements
reducedDims(s_sce) <- reducedDims(s_sce)[c("PCA","TSNE")]
colnames(reducedDim(s_sce, "TSNE")) <- c("TSNE1", "TSNE2")
colnames(reducedDim(s_sce, "PCA")) <- sub("_", "", 
                                          colnames(reducedDim(s_sce, "PCA")))

# Format row and column data
colnames(s_sce) <- NULL
rowData(s_sce)$geneID <- rownames(s_sce)

# Retain only raw counts
s_sce@assays@data@listData$logcounts <- NULL
s_sce@assays@data@listData$scaledata <- NULL

# Add Cell Ontology labels provided by authors
CO <- read.csv("~/MouseAgingData_data/2019nn_co.csv")

# Relabel csv column names. CO class and id are switched
colnames(CO) <- c("cell_type", "cell_ontology_class", "cell_ontology_id")

# Retain original cell_type levels
cell_levels <- levels(colData(s_sce)$cell_type)

# left join CO terms
colData(s_sce) <- DataFrame(colData(s_sce) %>% as_tibble() %>% 
                                left_join(CO, by = "cell_type"))

# Set levels for cell_type, cell_ontology_id, and cell_ontology_class
# Reorder levels based on aging biology and authors' study
colData(s_sce)$cell_type <- factor(colData(s_sce)$cell_type, levels = cell_levels)
levels(colData(s_sce)$cell_type) # check

colData(s_sce)$cell_ontology_class <- factor(colData(s_sce)$cell_ontology_class, 
                                             levels = unique(CO$cell_ontology_class), 
                                             exclude = NULL)

colData(s_sce)$cell_ontology_id <- factor(colData(s_sce)$cell_ontology_id, 
                                          levels = unique(CO$cell_ontology_id), 
                                          exclude = NULL)

dim(s_sce)
# should be 14699 37089
```

# Save `SingleCellExperiment` object

```{r}
saveRDS(s_sce, "brain10x_2019NN.rds")
```

# References

Ximerakis, M., Lipnick, S.L., Innes, B.T. et al. Single-cell transcriptomic
profiling of the aging mouse brain. *Nat Neurosci* 22, 1696–1708 (2019).
<DOI:https://doi.org/10.1038/s41593-019-0491-3>.

# Session info

```{r}
sessionInfo()
```